Detecting social media events through tweet content analysis

### 2020-05-06 Rmd/jse-2019-ms.Rmd
incorporated EACD suggestions

---
title: 'Detect social media events through tweet content analysis'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

Twitter has profoundly changed how we communicate. 
In only 280 characters, users speak their piece on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for text 
analysis now enable mentors to use tweets as pedagogical 
tools in guiding undergraduate research projects.

Some social media data, including tweets from Twitter, is
available through website application product interfaces (APIs). Twitter 
shares, via a streaming API, a sample of approximately one percent of all 
tweets during an API query time period [@tweet_stream]. Any Twitter user can 
access this one percent sample. For a price, researchers can get access to larger 
tweet samples.

With large collections of tweets, researchers have studied
inference of relationships and social networks among Twitter users [@lin2011joint]; 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a collection of tweets over time by repeated querying of the Twitter streaming API.

Our collection of tweets over time served as a valuable resource in our mentoring 
of undergraduate data science research. @nolan2010computing argue for students to 
work with real data. 
Working with real data allows students to develop skill not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, and using data 
from multiple file formats. Collaboratively asking and addressing novel questions with 
our collection of tweets gave mentored students 
opportunities to develop skills in all of these areas.

Mentoring in the work place and in higher education has many 
benefits, including improving 1) students' development as thinkers and scholars, 2) confidence in their 
own abilities, 3) integration into the campus community,
and 4) interest in graduate training 
[@baker2010beyond;@higgins2001reconceptualizing].

While our tweet collection enables us to address many possible research questions, the 
content of tweets over time particularly intrigued us. We 
hypothesized that high-profile social media events would generate lots of 
tweets, and that we'd detect social media events through changes in tweet 
topic content over time.

We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring of undergraduate 
students.

@@

We sought to validate our hypothesis that we could detect a major social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
examine tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted latent Dirichlet allocation models for each of five 
distinct one-hour periods. The first period began approximately 48h 
before the Super Bowl halftime show. Subsequent time periods started at -24, 
0, +24, and +48 hours after the approximate start time of the halftime show.

>>>>

---
title: 'Detecting social media events through tweet content analysis'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

Twitter has profoundly changed how we communicate. 
In only 280 characters, users can instantly contribute to public conversations on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for text 
analysis now enable mentors to use tweets as contemporary pedagogical 
tools in guiding undergraduate research projects.

Some social media data, including tweets from Twitter, is
available through website application product interfaces (APIs). 
By way of a streaming API, Twitter shares a sample of approximately one percent of all 
tweets during an API query time period [@tweet_stream]. Any Twitter user can freely 
access this one percent sample, whereas access to a larger selection is available to researchers for a fee.

Using large collections of tweets, researchers have studied
inference of relationships and social networks among Twitter users [@lin2011joint]; 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a collection of tweets over time by repeated querying of the Twitter streaming API.

In line with recent calls for students to work with real data [@nolan2010computing],   our collection of tweets has served as a valuable resource in our mentoring 
of undergraduate data science research. 
Working with real data allows students to develop proficiency not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, and using data 
from multiple file formats. Collaboratively asking and addressing novel questions with 
our collection of tweets gave mentored students 
opportunities to develop competency in all of these areas.

Mentoring in the work place and in higher education has many 
benefits, including improving 1) students' development as thinkers and scholars, 2) confidence in their 
own abilities, 3) integration into the campus community,
and 4) interest in graduate training [@baker2010beyond;@higgins2001reconceptualizing]. 
We strived to develop trusting, mutually respectful mentoring relationships with 
our students while advising their senior projects. With input from us,
the students selected appealing research questions.

While our tweet collection enables us to address many possible research questions, the 
content of tweets over time particularly intrigued us. We 
hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we'd detect social media events through changes in tweet 
topic content over time. We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring of undergraduate 
students.

@@

We sought to validate our hypothesis that we could detect a major social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted latent Dirichlet allocation models for each of five 
distinct one-hour periods. The first period began approximately 48h 
before the Super Bowl halftime show. Subsequent time periods started at -24, 
0, +24, and +48 hours after the approximate start time of the halftime show.
